---
title: "tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
This is a brief PHONEMeS tutorial. It provides an example for an analysis with 
PHONEMeS starting from the results of an differential analysis of phosphorylation 
sites, in this particular case a top table results object from limma.
PHONEMeS is a method to model signaling networks based on untargeted phosphoproteomics
mass spectrometry data and kinase/phosphatase-substrate interactions. It identifies
a subset of interactions from a prior knowledge network that represent potential 
regulatory signaling pathways linking known or potentially deregulated kinases to 
downstream phorsphorylation sites.

# Getting set up
First, we load the PHONEMeS library. To use PHONEMeS, in particular the run_phonemes
function, an interactive version of IBM Cplex or CBC-COIN solver is required as 
the network optimizer. The IBM ILOG Cplex is freely available through Academic 
Initiative [here](https://www.ibm.com/products/ilog-cplex-optimization-studio). 
The [CBC](https://projects.coin-or.org/Cbc) solver is open source and freely available
for any user. Alternatively for smaller cases, users can rely on the freely available 
[lpSolve R-package](https://cran.r-project.org/web/packages/lpSolve/index.html). 

Additionally, we load the [decoupleR](https://github.com/saezlab/decoupleR) library,
a recent development from the Saez-Rodriguez group which offers 12 different 
statistical approaches to perform functional analysis of omics data.

```{r setup}
library(PHONEMeS)
library(decoupleR)
library(tidyverse)
```

We then take a look at the tutorialData inside PHONEMeS. It consists of a top 
table results object from limma coming from phosphoproteomics. This contains the 
information about phosphorylation sites that are differntially abundant between 
two conditions. 

```{r}
head(tutorialData)
```

# Input preparation
For the extraction of perturbed signaling networks using PHONEMeS, potentially 
deregulated phosphorylation sites and kinases need to be selected that will be 
connected based on a prior knowledge network. 

## Selection of deregulated phosphorylation sites
Deregulated phosphorylation sites can be selected based on their changes detected
within the differential analysis. We choose T values for the selection over 
other metrices (p-values, logFC) as it combines the significance and magnitude 
of the change.
We then choose the top 10 deregulated phosphorylation sites that we will try to 
connect to upstream kinases. This is in arbitrary number and users should decide 
on a case to case basis what selection criteria to apply. The phosphorylation 
sites could also be chosen based on a chosen percentage or score cut-off.

```{r}
top_pps <- tutorialData %>% 
  dplyr::filter(ID %in% phonemesPKN$target) %>%
  dplyr::arrange(dplyr::desc(base::abs(t))) %>%
  dplyr::slice(1:10)

deregulated_pps <- top_pps$t
names(deregulated_pps) <- top_pps$ID
```


## Selection of deregulated kinases
Kinases that are deregulated can either be selected manually, e.g. if a specific 
kinase is known to be perturbed in an experiment, or based on changes in their 
activity. Therefore, we will estimate kinase activity changes using the decoupleR 
framework.


### Manual selection
For the manual selection, we create a names vector were kinases can be classified 
as eiter up-regulated (1) or down-regulated (-1). In this particular example we 
assume an up-regulation of AKT1.

```{r}
deregulated_kinases_man <- c(AKT1 = 1)
```

### Kinase activity estimation
DecoupleR is a collection of computational methods used for footprint-based analysis
that assume omics data as a signature of upstream biological activities. Here, we
can use the changes in phosphorylation sites as signatures for the activity of 
upstream kinases. The interactions between phosphorylation sites and kinases are 
again taken from the phonemesPKN.
To use the phonemesPKN for decoupleR we need to add a column for the likelihood 
of each interaction. Since our network is not weighted, we just add a 1 for all 
interactions.

```{r}
decoupler_network <- phonemesKSN %>% dplyr::rename("mor" = interaction) %>% tibble::add_column("likelihood" = 1)
```

As decoupleR functions expect a matrix as input, we are going to create a one-column
matrix from the tutorialData object. This one column matrix contains the vector 
of T values obtained from limma. We again choose the T values over other metrices.

```{r}
decoupler_input <- tutorialData %>% 
  dplyr::filter(ID %in% decoupler_network$target) %>%
  tibble::column_to_rownames("ID") %>% 
  dplyr::select(t)
```

Before running any method in decoupleR, we need intersect our prior knowledge 
network with the input matrix. With that we filter out regulons with less than 
5 target features.

```{r}
decoupler_network <- decoupleR::intersect_regulons(mat = decoupler_input, network = decoupler_network, 
                                                   .source = source, .target = target, minsize = 5)
```

Additionally, the overlapp of different regulons should be checked. If some kinases
have the same set of targets (correlatiom = 1), we only keep one of them.
```{r}
correlated_regulons <- decoupleR::check_corr(decoupler_network) %>% dplyr::filter(correlation == 1)
decoupler_network <- decoupler_network %>% dplyr::filter(!source %in% correlated_regulons$source.2)
```

We use mlm to estimate kinase activity but if the user would like to check other
available methods they can run show_methods(). All methods follow the same design
pattern and arguments.

```{r}
kinase_activity <- decoupleR::run_mlm(mat = decoupler_input, network = decoupler_network)
```

We then choose the top 5 deregulated kinases that we will try to connect to the 
downstream phosphorylation sites. This is again in arbitrary number and users 
should decide on a case to case basis what selection criteria to apply. Kinases 
could also be chosen based on a chosen percentage or score cut-off.
As before, kinases will eiter be classified as up-regulated (1) or down-regulated (-1).

```{r}
top_kinases <- kinase_activity %>% dplyr::arrange(dplyr::desc(base::abs(score))) %>% dplyr::slice(1:5)

deregulated_kinases <- top_kinases$score
names(deregulated_kinases) <- top_kinases$source

deregulated_kinases[deregulated_kinases > 0] <- 1
deregulated_kinases[deregulated_kinases < 0] <- -1
```

Moreover, we recommend the users to select kinases that do not appear to be regulated
based on their kinase activities. We assume that kinases with a score below 0.5 
are likely to not be deregulated between the conditions and with that do not explain
the observed changes. These kinases will be filtered out from our network within PHONEMeS.

```{r}
nc_kinases <- kinase_activity %>% dplyr::filter(base::abs(score) <= 0.5) %>% dplyr::pull(source)
```

# Run PHONEMeS
The deregulated phosphorylation sites and kinases can now be connected using PHONEMeS,
in specific the run_phonemes function. Within that function the prior knowledge 
network is pruned by removing nodes 50 steps upstream and downstream of measurements
and inputs. Remember that the path to an interactive version of IBM Cplex or CBC-COIN
solver is required.

This example should take about 1 minute to be solved by PHONEMeS.

```{r, results = "hide"}
phonemes_result <- PHONEMeS::run_phonemes(inputObj = deregulated_kinases, 
                                          measObj = deregulated_pps, 
                                          rmNodes = nc_kinases, 
                                          netObj = phonemesPKN,
                                          solverPath = "/Applications/CPLEX_Studio201/cplex/bin/x86-64_osx/cplex",  # Example path
                                          solver = "cplex")
```
The output of PHONEMeS contains the network, measurements and inputs used for the construction and the final results, including individual solutions and one summary solution (weightedSIF, nodesAttributes). These can be directly used to identify causal interactions between the perturbed nodes and the selected kinases. In addition to extracting direct information from the network, we can run different downstream analysis based on the necessities of each project, e.g. Pathway enrichment analysis.

```{r}
utils::str(phonemes_result,max.level = 2)
```


# Session Info
This tutorial was run on the date specified below.
```{r}
Sys.Date()
```

The sessionInfo() at run time was:
```{r}
sessionInfo()
```



